# (APPENDIX) Other stuff {-}
# QEPCAD-B output

## Quasi-affine { #sec:quasi-output }

The `QEPCAD-B` output from Example \@ref(exm:sphere) is shown below.

``` {r, child="test-cases/sphere.Rmd"}

```

## Monotone { #sec:monotone-output }

The `QEPCAD-B` output from Example \@ref(exm:qanm) is shown below.

``` {r, child="test-cases/qanm.Rmd"}

```

## Frontier Condition { #sec:frontier-output }

The `QEPCAD-B` output from Example \@ref(exm:fr-3-whitney-jhd) is shown below.

``` {r, child="test-cases/fr-3-whitney-jhd.Rmd"}

```

# More Test Cases { #sec:more-tests }

We first present a few variations of Example \@ref(exm:qanm), to test various situations. The following Example \@ref(exm:qanm-3-alg) requires an algebraic refinement point.

::: {.example #qanm-3-alg }
Let $n = 3$ and consider the 2-dimensional section cell
$$
C := \{ (x,y,z) \in \R^3 \mid 0 < x < 1, y > 0, y - x^2 + 2x < 1, z = x^2 + y^2 \}.
$$
:::

Below are the projection and refinement polynomials produced by `QEPCAD-B`. The resulting refinement point is also
shown, along with the original and refined cells.
\begin{align*}
\text{Projection polynomials:}\\
A_{1,1} & = input\\
        & = x\\
A_{1,2} & = fac2(J_{1,1}) = fac2(res(A_{2,1}|A_{2,2}))\\
        & = x - 1\\
M_{1,3} & = fac(K_{1,3}) = fac(input)\\
        & = 2 x^3 - 6 x^2 + 7 x - 2\\
M_{1,4} & = fac(K_{1,6}) = fac(input)\\
        & = 8 x - 9\\
A_{2,1} & = input\\
        & = y\\
A_{2,2} & = input\\
        & = y - x^2 + 2 x - 1\\
A_{3,1} & = input\\
        & = z - y^2 - x^2
\end{align*} \begin{align*}
\text{Refinement points:}\\
R_{1,1} & = input \text{ *** Refinement of cell (3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\alpha  & = \text{the unique root of } 2 x^3 - 6 x^2 + 7 x - 2 \text{ between } 1/4 \text{ and } 1/2\\
\text{Coordinate }1 & = \alpha \approx 0.4102\ldots\\
\text{Original cell: }
& C_{(3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 1/2 = 0.5000000000\\
\text{Coordinate }2 & = 1/8 = = 0.1250000000\\
    \text{Coordinate }3 & = 19/64 = 0.2968750000\\
\text{Refined cells: }
& C_{(3,3,2)}\\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 13/64 = 0.2031250000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
    \text{Coordinate }3 & = 425/4096 \approx 0.1037597656\ldots\\
& C_{(4,3,2)} \\
& \text{The sample point is in an EXTENDED representation.}\\
\alpha & = \text{the unique root of } 2 x^3 - 6 x^2 + 7 x - 2 \text{ between } 3/8 \text{ and } 7/16 \\
\text{Coordinate }1 & = \alpha \approx 0.4102454877\ldots\\
\text{Coordinate }2 & = 1/8 = 0.1250000000\\
\text{Coordinate }3 & = \text{the unique root of } \\
& 262144 x^3 - 536576 x^2 + 1654976 x - 287873 \text{ between } 0 \text{ and } 4\\
                    & \approx 0.1839263602\ldots\\
& C_{(5,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 1/2 = 0.5000000000\\
\text{Coordinate }2 & = 1/16 = 0.0625000000\\
\text{Coordinate }3 & = 65/256 = 0.2539062500
\end{align*}

Example \@ref(exm:qanm-4-rat), which follows, demonstrates what happens on sub-CADs above 0-dimensional cells.

::: {.example #qanm-4-rat }
Let $n = 4$ and consider the semialgebraic set defined by the QFF
$$
F := ( ( 4 w = 7 \lor w^2 = 36 ) \land x > 0 \land x < 1 \land y > 0 \land x + y < 1 \land z = w x^2 + y^2 ).
$$
It has three connected components, $C_1$, $C_2$ and $C_3$, each of which is a cylindrical section cell in a sub-CAD of $\R^3$ above a $0$-dimensional cell ($\mathbf{b}_1$, $\mathbf{b}_2$, $\mathbf{b}_3$ respectively).
\begin{align*}
\mathbf{b}_1 &= 7/4,& &C_1 &= \{ (x,y,z) \in \R^3 \mid x > 0 ,x < 1 ,y > 0 ,x + y < 1 ,z = 7/4 x^2 + y^2 \}.\\
\mathbf{b}_2 &= -6,& &C_3 &= \{ (x,y,z) \in \R^3 \mid x > 0 ,x < 1 ,y > 0 ,x + y < 1 ,z = -6 x^2 + y^2 \}.\\
\mathbf{b}_3 &= 6,& C_3 &= \{ (x,y,z) \in \R^3 \mid x > 0 ,x < 1 ,y > 0 ,x + y < 1 ,z = 6 x^2 + y^2 \}.
\end{align*}

$C_2$ is already monotone. This can be seen by examining the projections of its top and bottom onto $\opspan{x,z}$:
\begin{align*}
C_{2,B} &= \{ -1 < x < 1, z = -6x^2 \},\\
C_{2,T} &= \{ -1 < x < 1, z = -5x - 2x + 1 \}.
\end{align*}

$C_1$ and $C_3$ are of the same kind as the cell presented in Example \@ref(exm:qanm) and require refinement.
:::

Projection polynomials and refinement points output by `QEPCAD-B` are presented below.
\begin{align*}
\text{Projection polynomials:}\\
A_{1,1} & = input\\
        & = 4 w - 7\\
A_{1,2} & = input\\
        & = w - 6\\
A_{1,3} & = input\\
        & = w + 6\\
Q_{1,4} & = fac(Q_{2,2}) = fac(input)\\
        & = w\\
A_{2,1} & = input\\
        & = x\\
A_{2,2} & = fac(J_{2,1}) = fac(res(A_{3,1}|A_{3,2}))\\
        & = x - 1\\
M_{2,3} & = fac(K_{2,4}) = fac(input)\\
        & = 5 x + 1\\
M_{2,4} & = fac(K_{2,6}) = fac(input)\\
        & = 8 x - 1\\
M_{2,5} & = fac(K_{2,7}) = fac(input)\\
        & = 7 x - 1\\
A_{3,1} & = input\\
        & = y\\
A_{3,2} & = input\\
        & = y + x - 1
\end{align*}
\begin{align*}
\text{Refinement points:}\\
R_{2,2} & = input \text{ *** Refinement of cell (8,3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 6 = 6.0000\\
\text{Coordinate }2 & = 1/7 \approx 0.1429\ldots\\
R_{2,1} & = input \text{ *** Refinement of cell (6,3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 7/4 = 1.7500\\
\text{Coordinate }2 & = 1/8 = 0.1250
\end{align*} \begin{align*}
\text{Original cell: } C_1
& C_{(6,3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 7/4 = 1.7500000000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 11/64 = 0.1718750000\\
\text{Refined cells: }
& C_{(6,3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 7/4 = 1.7500000000\\
\text{Coordinate }2 & = 1/16 = 0.0625000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 &= 71/1024 = 0.0693359375\\
& C_{(6,4,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 7/4 = 1.7500000000\\
\text{Coordinate }2 & = 1/8 = 0.1250000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 23/256 = 0.0898437500\\
& C_{(6,5,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 7/4 = 1.7500000000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 11/64 = 0.1718750000\\
\end{align*} \begin{align*}
\text{Original cell: }
C_3 & C_{(8,3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 6 = 6.0000000000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 &= 7/16 = 0.4375000000\\
\text{Refined cells: }
& C_{(8,3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 6 = 6.0000000000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 7/16 = 0.4375000000\\
& C_{(8,4,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 6 = 6.0000000000\\
\text{Coordinate }2 & = 1/7 \approx 0.1428571429\ldots \\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 145/784 \approx 0.1849489796\ldots\\
& C_{(8,3,3,2)} \\
& \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 6 = 6.0000000000\\
\text{Coordinate }2 & = 1/4 = 0.2500000000\\
\text{Coordinate }3 & = 1/4 = 0.2500000000\\
\text{Coordinate }4 & = 7/16 = 0.4375000000\\
\end{align*}

The following example requires a rational refinement above an algebraic zero-cell.

::: {.example #qanm-4-above-alg }
Let $n = 4$ and consider the section cell
$$
C := \{ (w,x,y,z) \in \R^4 \mid w = +\sqrt{3}, 0 < x < 1, y > 0, x + y < 1, z = 3x^2 + 2y^2 \}.
$$
Note thet $w = +\sqrt{3}$ is represented as the QFF $\{ w > 0, w^2 = 3 \}$.
:::

This cell lies in the sub-cad of $\R^3$ above $\sqrt{2}$. `QEPCAD` output is presented below. Note that the sample points remain in extended representation, which is how they were generated in the lifting phase. Polynomials and refinement points are as follows:
\begin{align*}
\text{Projection polynomials:}\\
A_{1,1} & = input\\
        & = w\\
A_{1,2} & = input\\
        & = w^2 - 2\\
A_{2,1} & = input\\
        & = x\\
A_{2,2} & = fac(J_{2,1}) = fac(res(A_{3,1}|A_{3,2}))\\
        & = x - 1\\
M_{2,3} & = fac(K_{2,3}) = fac(input)\\
        & = 5 x - 2\\
A_{3,1} & = input\\
        & = y\\
A_{3,2} & = input\\
        & = y + x - 1\\
A_{4,1} & = input\\
        & = z - 2 y^2 - 3 x^2\\
\text{Refinement points:}\\
R_{2,1} & = input \text{ *** Refinement of cell (6,3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\alpha  & = \text{the unique root of } x^2 - 2 \\
        & \text{ between } 181/128 \text{ and } 1449/1024\\
\text{Coordinate }1 & = \alpha \approx 1.4142\ldots\\
\text{Coordinate }2 & = 2/5 = 0.4000\\
\end{align*}

A rational refinement of the cell with positional index $(6,3)$ takes place. The sample points of new cells are recomputed and are stored in the extended representation, as with the original cell.

We now show a similar example, requiring an algebraic refinement.

::: {.example #qanm-4-alg }
Let $n = 4$ and consider the section cell
$$
C := \{ (w,x,y,z) \in \R^4 \mid w = + \sqrt{3}, 0 < x < 1, y > 0, y < (x-1)^2, z = 3x^2 + 2y^2 \}.
$$
Note that $w = + \sqrt{3}$ is represented by the QFF $\{ w > 0, w^2 = 3 \}$.
:::

The following polynomials and one refinement point are output by `QEPCAD-B`. The refinement point is stored in the extended represeentation.
\begin{align*}
\text{Projection polynomials:}\\
A_{1,1} & = input\\
        & = w\\
A_{1,2} & = input\\
        & = w^2 - 3\\
A_{2,1} & = input\\
        & = x\\
A_{2,2} & = fac2(J_{2,1}) = fac2(res(A_{3,1}|A_{3,2}))\\
        & = x - 1\\
M_{2,3} & = fac(K_{2,3}) = fac(input)\\
        & = 2 x^3 - 6 x^2 + 7 x - 2\\
M_{2,4} & = fac(K_{2,6}) = fac(input)\\
        & = 8 x - 9\\
A_{3,1} & = input\\
        & = y\\
A_{3,2} & = input\\
        & = y - x^2 + 2 x - 1\\
A_{4,1} & = input\\
        & = z - y^2 - x^2\\
\text{Refinement points:}\\
R_{2,1} & = input \text{ *** Refinement of cell (6,3) ***}\\
        & = \text{The sample point is in an EXTENDED representation.}\\
\alpha  & = \text{the unique root of } x^2 - 3 \text{ between } 1773/1024 \text{ and } 887/512\\
\text{Coordinate }1 & = \alpha \approx 1.7321\ldots\\
\text{Coordinate }2 & = \text{the unique root of } 2 x^3 - 6 x^2 + 7 x - 2 \text{ between } 1/4 \text{ and } 1/2\\
& \approx 0.4102\ldots\\
\end{align*}

An algebraic refinement is performed. In order to recompute sample points of the refined cells, the sample points of the cells contained in $C_{(6,3)}$ (in the decomposition induced on $\R^2$) are converted back into primitive representation. For example, a new cell $C_{(6,4,3,2)}$ has the sample point
\begin{align*}
& \text{The sample point is in an EXTENDED representation.}\\
\alpha & = \text{the unique root of } 4 x^6 - 24 x^5 + 28 x^4 + 52 x^3 - 35 x^2 - 64 x - 107\\
       & \text{ between } 17/8 \text{ and } 69/32\\
\text{Coordinate }1 & = \frac{300}{8777} \alpha^5 - \frac{1518}{8777} \alpha^4 + \frac{322}{8777} \alpha^3 + \frac{5607}{8777} \alpha^2 + \frac{6683}{8777} \alpha - \frac{9577}{8777}\\
& \approx 1.7320508076\ldots\\
\text{Coordinate }2 & = -\frac{300}{8777} \alpha^5 + \frac{1518}{8777} \alpha^4 - \frac{322}{8777} \alpha^3 - \frac{5607}{8777} \alpha^2 + \frac{2094}{8777} \alpha + \frac{9577}{8777} \\
& \approx 0.4102454877\ldots\\
\text{Coordinate }3 & = 1/8 = 0.1250000000\\
\text{Coordinate }4 & = \text{the unique root of } 262144 x^3 - 536576 x^2 + 1654976 x - 287873 \\
                    & \text{ between } 0 \text{ and } 4 \approx 0.1839263602\ldots
\end{align*}

Now consider a $(1,1,0,0)$-cell in $\R^4$ requiring two refinements.

::: {.example #qanm-4-multi }
Let $n = 4$ and consider the section cell
$$
C := \{ (w,x,y,z) \in \R^4 \mid  0 < w < 1, x > 0, w + x < 1, y = w^2 + x^2, z = 2 w^2 + x^2 \}.
$$
:::

Polynomials and refinement points as follows:
\begin{align*}
\text{Projection polynomials:}\\
A_{1,1} & = input\\
        & = w\\
A_{1,2} & = fac(J_{1,1}) = fac(res(A_{2,1}|A_{2,2}))\\
        & = w - 1\\
M_{1,3} & = fac(K_{1,4}) = fac(input)\\
        & = 3 w - 1\\
M_{1,4} & = fac(K_{1,5}) = fac(input)\\
        & = 2 w - 1\\
A_{2,1} & = input\\
        & = x\\
A_{2,2} & = input\\
        & = x + w - 1\\
A_{3,1} & = input\\
        & = y - x^2 - w^2\\
A_{4,1} & = input\\
        & = z - x^2 - 2 w^2
\end{align*} \begin{align*}
\text{Refinement points:}\\
R_{1,2} & = input \text{ *** Refinement of cell (3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 1/2 = 0.5000\\
R_{1,1} & = input \text{ *** Refinement of cell (3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 1/3 \approx 0.3333\ldots\\
\end{align*}

The following example, in $\R^5$ presents a $(1,0,1,0,0)$-cell, which requires two algebraic refinements of the CAD induced on $\R^1$. When running this test, `SACLIB` ran out of memory, so the number of available cells was doubled.

::: {.example #qanm-5-101 }
Let $n = 5$ and consider the $(1,0,1,0,0)$-cell
$$
C = \{ 0 < w < 1, x = 1 - w, y > 0, w + y < 1, z = w^2 + x y^2, t = 3 w^2 + x^2 y^2 \}.
$$
The CAD will be constructed with variable ordering $w \prec x \prec y \prec z \prec t$.
:::

The projection polynomials produced by `QEPCAD-B` are as follows:
\begin{align*}
A_{1,1} & = fac(J_{2,2}) = fac(ldcf(red(Q_{3,4})) = res(A_{3,1}|Q_{3,4})) = fac(J_{2,3})\\
        & = fac(ldcf(red(Q_{3,5})) = res(A_{3,1}|Q_{3,5})) = fac(J_{2,4}) = fac(dis(Q_{3,3}))\\
        & = fac(J_{2,5}) = fac(dis(Q_{3,4})) = fac(J_{2,6}) = fac(dis(Q_{3,5})) = fac(J_{2,8})\\
        & = fac(res(A_{3,1}|Q_{3,3})) = fac2(J_{2,12}) = fac2(res(Q_{3,3}|Q_{3,4})) = fac2(J_{2,13})\\
        & = fac2(res(Q_{3,3}|Q_{3,5})) = fac2(J_{2,14}) = fac2(res(Q_{3,4}|Q_{3,5})) = fac(J_{1,7})\\
        & = fac(res(S_{2,2}|P_{2,4})) = fac(J_{1,8}) = fac(res(S_{2,2}|P_{2,5})) = fac(J_{1,11})\\
        & = fac(res(P_{2,4}|P_{2,5}))\\
        & = w\\
A_{1,2} & = fac(J_{2,1}) = fac(res(A_{3,1}|A_{3,2})) = fac(J_{1,1}) = fac(res(A_{2,1}|S_{2,2}))\\
        & = fac(J_{2,7}) = fac(res(A_{3,1}|A_{3,2})) = fac2(J_{1,1}) = fac2(ldcf(P_{2,4}))\\
        & = fac(J_{1,2}) = fac(res(A_{2,1}|S_{2,2})) = fac2(J_{1,10}) = fac2(res(P_{2,3}|P_{2,5}))\\
        & = w - 1\\
P_{1,3} & = fac(J_{2,9}) = fac(res(A_{3,2}|Q_{3,3})) = fac(J_{1,11}) = fac(res(P_{2,4}|P_{2,5}))\\
        & = fac(J_{1,12}) = fac(res(P_{2,4}|P_{2,6}) = res(P_{2,5}|P_{2,6}))\\
        & = w^2 - 4 w + 1\\
P_{1,4} & = fac(J_{1,3}) = fac(res(A_{2,1}|P_{2,3}))\\
        & = w + 2\\
P_{1,5} & = fac(J_{1,4}) = fac(res(A_{2,1}|P_{2,4}))\\
        & = w^3 - 3 w^2 + 6 w - 1\\
P_{1,6} & = fac(J_{1,5}) = fac(res(A_{2,1}|P_{2,5}))\\
        & = w^3 - w^2 + 7 w - 1
P_{1,7} & = fac(J_{1,6}) = fac(res(A_{2,1}|P_{2,6}))\\
        & = 2 w + 1\\
P_{1,8} & = fac(J_{1,9}) = fac(res(P_{2,3}|P_{2,4}))\\
        & = w^2 - 3 w + 1\\
M_{1,9} & = fac(K_{1,15}) = fac(input)\\
        & = w + 1\\
M_{1,10} &= fac(K_{1,16}) = fac(input)\\
         &= 2 w^3 - 6 w^2 + 9 w - 2\\
M_{1,11} &= fac(K_{1,17}) = fac(input)\\
         &= w + 6\\
M_{1,12} &= fac(K_{1,19}) = fac(input)\\
         &= 3176 w - 4353\\
M_{1,13} &= fac(K_{1,20}) = fac(input)\\
         &= 3 w^2 - 8 w + 3
\end{align*} \begin{align*}
A_{2,1} & = input\\
        & = x + w - 1\\
S_{2,2} & = fac(input) = fac(J_{2,1}) = fac(ldcf(Q_{3,4}) = ldcf(Q_{3,5}))\\
        & = fac(J_{2,5}) = fac(dis(Q_{3,4})) = fac(J_{2,6})\\
        & = fac(dis(Q_{3,5})) = fac2(J_{2,14}) = fac2(res(Q_{3,4}|Q_{3,5}))\\
        & = x\\
P_{2,3} & = fac(J_{2,3}) = fac(ldcf(red(Q_{3,5})) = res(A_{3,1}|Q_{3,5}))\\
        & = fac(J_{2,6}) = fac(dis(Q_{3,5}))\\
        & = x - 3\\
P_{2,4} & = fac(J_{2,10}) = fac(res(A_{3,2}|Q_{3,4}))\\
        & = w^2 x - 2 w x + x - 3 w\\
P_{2,5} & = fac(J_{2,11}) = fac(res(A_{3,2}|Q_{3,5}))\\
        & = w^2 x + x - 6 w\\
P_{2,6} & = fac2(J_{2,12}) = fac2(res(Q_{3,3}|Q_{3,4}))\\
        & = fac2(J_{2,13}) = fac2(res(Q_{3,3}|Q_{3,5})) = fac2(J_{2,14}) = fac2(res(Q_{3,4}|Q_{3,5}))\\
        & = 2 x - 3
\end{align*} \begin{align*}
A_{3,1} & = input\\
        & = y\\
A_{3,2} & = input\\
        & = y + w - 1\\
Q_{3,3} & = input\\
        & = y^2 - 2 w\\
Q_{3,4} & = input\\
        & = x y^2 - 3 w\\
Q_{3,5} & = input\\
        & = x y^2 + 2 w x - 6 w\\
A_{4,1} & = input\\
        & = z - x y^2 - w^2\\
A_{5,1} & = input\\
        & = t - x^2 y^2 - 3 w^2
\end{align*}
Note the large number of polynomials. In particular, those indexed by $Q$ and $S$, resulting from the quasi-affine step of the algorithm.

The refinement polynomials (labelled by $M$) define two refinement points in the true cell $C_{(7)}$:
\begin{align*}
R_{1,1} & = input \text{ *** Refinement of cell (7) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\alpha_{1}  & = \text{the unique root of } 2 x^3 - 6 x^2 + 9 x - 2 \text{ between } 1/4 \text{ and } 5/16\\
\text{Coordinate }1 & = \alpha \approx 0.2649\ldots\\
R_{1,2} & = input \text{ *** Refinement of cell (11) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\alpha_{2}  & = \text{the unique root of } 3 x^2 - 8 x + 3 \text{ between } 3/8 \text{ and } 1/2\\
\text{Coordinate }1 & = \alpha \approx 0.4514\ldots
\end{align*}

::: {.example #fr-4-whitney-multi }
Let $n = 4$ and consider the semialgebraic set defined by the QFF, containing polynomials in $Z[t,x,y,z]$,
$$
F = \{ (t = 0 \lor t = 2) \land 0 < x < 1 \land -x < y < x \land y^2 - x^2 (z - t) = 0 \}.
$$
:::

The QFF defines two cylindrical cells in $\R^3$ (equipped with $(x,y,z)$), similar to those defined in Example \@ref(exm:whitney-jhd).
The following refinement points are obtained.
\begin{align*}
R_{4,2} & = input \text{ *** Refinement of cell (6,2,2,5) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 2 = 2.0000\\
\text{Coordinate }2 & = 0 = 0.0000\\
\text{Coordinate }3 & = 0 = 0.0000\\
\text{Coordinate }4 & = 3 = 3.0000\\
R_{4,1} & = input \text{ *** Refinement of cell (4,2,2,3) ***}\\
        & = \text{The sample point is in a PRIMITIVE representation.}\\
\text{Coordinate }1 & = 0 = 0.0000\\
\text{Coordinate }2 & = 0 = 0.0000\\
\text{Coordinate }3 & = 0 = 0.0000\\
\text{Coordinate }4 & = 1 = 1.0000
\end{align*}

A similar example is presented below.

::: {.example #fr-4-whitney-alg }
Let $n = 4$ and consider the twe $2$-diemnsional subsets of the Whitney whitney umbrella defined by the QFF
$$
F = \{ (t,x,y,z) \in \R^4 \mid t^2 = 0 \land 0 < x < 1 \land -x < y < x \land y^2 - x^2 (z - t) = 0 \}.
$$
:::

This time, algebraic refinements are required:
\begin{align*}
R_{4,3} & = input \text{ *** Refinement of cell (8,2,2,1) ***}\\
        & = \text{The sample point is in an EXTENDED representation.}\\
\alpha  & = \text{the unique root of } x^2 - 2 \text{ between } 181/128 \text{ and } 1449/1024\\
\text{Coordinate }1 & = \alpha \approx 1.4142\ldots\\
\text{Coordinate }2 & = 0 = 0.0000\\
\text{Coordinate }3 & = 0 = 0.0000\\
\text{Coordinate }4 & = \text{the unique root of } x^2 - 2 x - 1 \text{ between } -1/2 \text{ and } -1/4\\
                    & \approx -0.4142\ldots
\end{align*} \begin{align*}
R_{4,2} & = input \text{ *** Refinement of cell (2,2,2,3) ***}\\
        & = \text{The sample point is in an EXTENDED representation.}\\
\alpha  & = \text{the unique root of } x^2 - 2 \text{ between } -1449/1024 \text{ and } -181/128\\
\text{Coordinate }1 & = \alpha \approx -1.4142\ldots\\
\text{Coordinate }2 & = 0 = 0.0000\\
\text{Coordinate }3 & = 0 = 0.0000\\
\text{Coordinate }4 & = \text{the unique root of } x^2 - 2 x - 1 \text{ between } 0 \text{ and } 4\\
                    & \approx 2.4142\ldots\\
R_{4,1} & = input \text{ *** Refinement of cell (2,2,2,3) ***}\\
        & = \text{The sample point is in an EXTENDED representation.}\\
\alpha  & = \text{the unique root of } x^2 - 2 \text{ between } -1449/1024 \text{ and } -181/128\\
\text{Coordinate }1 & = \alpha\approx -1.4142\ldots\\
\text{Coordinate }2 & = 0 = 0.0000\\
\text{Coordinate }3 & = 0 = 0.0000\\
\text{Coordinate }4 & = \text{the unique root of } x^2 - 2 x - 1 \text{ between } -1/2 \text{ and } -1/4\\
                    & \approx -0.4142\ldots
\end{align*}

This will result in refinements of $1$-dimensional sector cells above, respectively, the cells with positional indices $(2,2,2)$ and $(8,2,2)$.

